Extreme heterogeneity of influenza virus infection in single cells

### 2018-01-13 paper/ResponseToReviews.tex

{\color{black}
This typo has been fixed.
}

- Lines 330-4: This statement about a confluence of factors leading to extreme cell-to-cell heterogeneity should be reworded since the data here actually doesn't show a clear role for innate immunity in promoting heterogeneity and there is no significant evaluation of the effect of host cell state on heterogeneity.

{\color{red} needs to be re-worded}

- The ``replicate'' 8-hr sample isn't really a replicate since a different MOI was used. Also, the purpose of the differences in treatment for this sample compared with the 8-hr sample are not well explained.

{\color{black}
This point is correct.
The second 8-hour sample (8hr-2) was done slightly differently, since the cells were not washed and the infecting MOI was slightly less.
We have clarified the text to emphasize that the 8hr-2 sample is \emph{not} simply a replicate.

The purpose of the 8hr-2 sample was to assess whether timing of infection was a major contributor to heterogeneity.
This sample is useful for looking at that question because the omission of the washing step means that infection for this sample can in principle occur at any time over the experimental window.
This is now clearly explained in the text.
For more details, see our response to MAJOR COMMENT \#1 above, which deals with the question of the role of infection timing / synchronization.
}

@@

-Figure 4C: the threshold line appears to be a bit to the right of the global maximum - going through the description, it was not clear why that should be the case.

{\color{black} 
This is a good observation. 
We were determining the threshold as the maximum of a loess fit to the data. 
Previously, the loess fitting parameters caused the fit line to overshoot, putting the threshold a bit to the right.
We have adjusted span parameter in the loess fitting, which fixes the overshoot.
The revised Figure 4C shows the loess fit in orange as well as the threshold in dotted green -- and the threshold now goes right through the maximum.
}

-Why was media change skipped in the replicate experiment?

{\color{black} 
This has been clarified in the text.
See the response to MINOR COMMENT three before this one, and to MAJOR COMMENT \#1.}

### 2018-01-13 paper/paper.tex
put pathway enrichment in sorted CSV rather than unsorted TSV
clarified that 8hr-2 is not really a replicate, but done slightly differently

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, including a replicate a 8-hours.
We replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
The exception was the replicate 8-hour sample, for which we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B).

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
\jdbcomment{add gray arrow wedge}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.tsv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

@@

To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots of the clarified viral supernatant at -80 \si{\degreeCelsius }. 
We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus replicate for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours.

@@

All time points except for the 8-hour replicate were prepared on the same day.
For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
A single well of cells was trypsinized and counted prior to initiation of the experiment for the purposes of calculating MOI.
Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
At the initiation of our experiment, the wells for all time points were changed from D10 to influenza growth media.
Cells were then infected with 0.3 HA-expressing units of virus per cell (a determined by flow cytometry).
The infections were performed in order of time point: first the 10-hour time point, then the 8-hour, and then the 6-hour time point.
At one hour after infection, the media for each time point was changed to fresh influenza growth media.
Note that the HA-expressing units were calculated without this additional washing step, and so likely represent an overestimate of our final infectious dose (consistent with the fact that fewer than 30\% of cells appear infected in the single-cell sequencing data).
All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media .
For our replicate 8-hour time point, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the manufacturer's provided by 10x Genomics for the Chromium Single Cell platform.
All subsequent steps through library preparation followed the manufacturer's protocol.
Samples were sequenced on an Illumina HiSeq. 
The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Upload in progress.}

@@

In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}).
We then determined the threshold fraction of viral mRNA at which the mean purity no longer increased as a function of the amount of viral mRNA.
This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
As is apparent from Figure~\ref{fig:viralbarcodes}B, only the 10-hour sample and the 8-hour-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
The likely reason is that these samples have more total viral mRNA (and so there is more available mRNA to be acquired from lysed cells); in addition, there is always some experimental variability in the amount of cell lysis during the 10X sequencing process, and these samples may simply have the most.
So the above threshold procedure is appropriate for those two samples.
For the other samples, we simply set a minimum threshold of requiring at least a fraction 2$\times$$10^{-4}$ reads to come from viral mRNA as explained in the legend to Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
The thresholds for each sample are shown in Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
For subsequent analyses, we retained all infected cells and a subsample of uninfected cells (the greater of 50 or the number of infected cells for that sample).
The rationale for subsampling the uninfected cell is that the vast majority of cells are uninfected, and we did not want these cells to completely dominate the downstream analyses.
Cells were classified as co-infected if both viral variants had an RNA level that exceeded the threshold, and if the minor variant contributed at least 5\% of the viral mRNA.

>>>>

\subsection{Single cells show an extremely wide range of expression of viral mRNA.}
We infected A549 cells at low MOI with a mixture of the wild-type and synonymously barcoded viruses, and collected cells for sequencing at 6, 8, and 10 hours post-infection, performing two slightly different variants of the experiment for the 8-hour timepoint.
For most of the samples, we replaced the infection inoculum with fresh media at one-hour post-infection, thereby ensuring that most infection was initiated during a narrow time window.
However, for the second 8-hour sample (which we denote as ``8hr-2'' in the figures), we did \emph{not} perform this media change and instead left the cells in the original infection inoculum.
We recovered between 3,000 and 4,000 cells for each sample (Figure~\ref{fig:cells}A). 
As expected for a low-MOI infection, most cells expressed little or no viral mRNA (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
Also as expected, the amount of viral mRNA per cell among infected cells increased over time (Figure~\ref{fig:cells}B and Figure~\ref{fig:cells}-Figure~supplement~\ref{figsupp:viral_mRNA_cumulfrac}).
But what was most notable was how widely the number of viral mRNA molecules varied among infected cells.
While the fraction of mRNA derived from virus was $<$0.1\% for most cells, viral mRNA constituted half the transcriptome in a few cells at 8 and 10 hours (Figure~\ref{fig:cells}B).

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap.pdf}}
\caption{\label{fig:cellulargenes}
\jdbcomment{add gray arrow wedge}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.csv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

@@

To generate viruses from these plasmids, we transfected an equimolar mix of all eight plasmids into cocultures of 293T and MDCK-SIAT1 cells seeded at a ratio of 8:1. 
At 24 hours post-transfection, we changed media from D10 to influenza growth media (Opti-MEM supplemented with 0.01\% heat-inactivated FBS, 0.3\% BSA, 100 U of penicillin/ml, 100  \si{\micro}g of streptomycin/ml, and 100 \si{\micro}g of calcium chloride/ml).
At 48 hours post-transfection we harvested the virus-containing supernatant, pelletted cellular material by centrifugation at 300 x g's for 4 minutes, and stored aliquots of the clarified viral supernatant at -80 \si{\degreeCelsius }. 
We then titered thawed aliquots of viral by TCID50 on MDCK-SIAT1 cells, computing titers via the formula of \citet{reed1938simple}.
To generate our ``high-purity'' stocks of viruses for the single-cell sequencing experiments, we then infected MDCK-SIAT1 cells at an MOI of 0.01, and let the virus grow for 36 hours prior to harvesting aliquots that were again clarified by low-speed centrifugation, aliquoted, stored at -80  \si{\degreeCelsius }, and titered by TCID50.
The high-MOI passage (high-defective particle) stock used in Figure~\ref{fig:viruspopulations} was generated by instead passaging in MDCK-SIAT1 cells twice at an MOI of 1 for 48 hours.

@@

All time points except for the second 8-hour sample (8hr-2) were prepared on the same day.
For the infections, A549 cells were seeded in a 6-well plate, with two wells per time point. 
A single well of cells was trypsinized and counted prior to initiation of the experiment for the purposes of calculating MOI.
Wild-type and synonymously barcoded virus were mixed to an estimated ratio of 1:1 based on prior, exploratory, single-cell analyses (data not shown). 
At the initiation of our experiment, the wells for all time points were changed from D10 to influenza growth media.
Cells were then infected with 0.3 HA-expressing units of virus per cell (a determined by flow cytometry).
The infections were performed in order of time point: first the 10-hour time point, then the 8-hour, and then the 6-hour time point.
At one hour after infection, the media for each time point was changed to fresh influenza growth media.
Note that the HA-expressing units were calculated without this additional washing step, and so likely represent an overestimate of our final infectious dose (consistent with the fact that fewer than 30\% of cells appear infected in the single-cell sequencing data).
All cells were then harvested for single-cell analysis concurrently -- ensuring all had spent equivalent time in changed media .
For 8hr-2 sample, cells were infected as above except that the cells were infected at 0.1 HA-expressing units of virus per cell but no wash step was performed, and the sample was prepared on a different day.
After harvest, cells were counted using disposable hemocytometers and diluted to equivalent concentrations with an intended capture of 3000 cells/sample following the manufacturer's provided by 10x Genomics for the Chromium Single Cell platform.
All subsequent steps through library preparation followed the manufacturer's protocol.
Samples were sequenced on an Illumina HiSeq. 
The deep sequencing data are available on the Sequence Read Archive under accession \jdbcomment{Upload in progress.}

@@

In order to determine an appropriate cutoff for how many reads a cell needed to contain in order to be classified as infected, we calculated the mean viral barcode purity across all cells that contained at least a given fraction of viral mRNA and had multiple viral reads that could be assigned a barcode (Figure~\ref{fig:viralbarcodes}B,C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}).
We then determined the threshold fraction of viral mRNA at which the mean purity no longer increased as a function of the amount of viral mRNA.
This threshold represents the point at which we have effectively eliminated cells that have low barcode purity simply due to lysis-acquired reads sampled randomly from both viral barcodes.
As is apparent from Figure~\ref{fig:viralbarcodes}B, only the 10-hour sample and the 8hr-2 sample have the excess of mixed barcodes among cells with low amounts of viral mRNA.
The likely reason is that these samples have more total viral mRNA (and so there is more available mRNA to be acquired from lysed cells); in addition, there is always some experimental variability in the amount of cell lysis during the 10X sequencing process, and these samples may simply have the most.
So the above threshold procedure is appropriate for those two samples.
For the other samples, we simply set a minimum threshold of requiring at least a fraction 2$\times$$10^{-4}$ reads to come from viral mRNA as explained in the legend to Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
The thresholds for each sample are shown in Figure~\ref{fig:viralbarcodes}C and Figure~\ref{fig:viralbarcodes}-Figure~supplement~\ref{figsupp:thresholds}.
This procedure is expected to be conservative, and may miss some truly infected cells with very low amounts of viral mRNA.
For subsequent analyses, we retained all infected cells and a subsample of uninfected cells (the greater of 50 or the number of infected cells for that sample).
The rationale for subsampling the uninfected cell is that the vast majority of cells are uninfected, and we did not want these cells to completely dominate the downstream analyses.
Cells were classified as co-infected if both viral variants had an RNA level that exceeded the threshold, and if the minor variant contributed at least 5\% of the viral mRNA.
